---
title: "Culex Larvae Colorimetry"
author: "Helle Aronson - Clément Vinauger"
date: "2025-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***NOTE***: This notebook produces the dotplots with boxplots and violin plots for each nutrient type presented in Figure 5 of Gregory *et al.*, Stage-specific divergence in larval thermal tolerance among populations of the *Culex pipiens* Assemblage. Journal of Medical Entomology, accepted for publication on June 18, 2026.

Execution of the code requires the raw data contained in the files: 
- "Culex_Protein_July2025.xlsx"
- "Culex_Glycogen_July2025.xlsx"
- "Culex_Total Lipids_July2025.xlsx"
- "Culex_Triglycerides_July2025.xlsx"


```{r, include=FALSE}
#Load Packages#

library(tidyverse)
library(ggpubr)
library(ggplot2)

#Data Set working directory#
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

```


```{r, echo = FALSE}
# Assign Colors for Populations using the color code from Ben

##Assign Colors

color.FL_MD <- "#DC9F1F" 
color.ILAG_MD <- "#6FB4E7"
color.ILBG_MD <- "#C27CA6"
color.ME_MD <- "#3373B0"
color.TX_MD <- "#ECE24E"



color_order<- c("#DC9F1F","#6FB4E7", "#C27CA6", "#3373B0","#ECE24E")


```

# Total Protein


```{r, echo = FALSE}
#### Protein ########################################################
######################################################################

##Load data
pf <- readxl::read_xlsx("Culex_Protein_July2025.xlsx")

##Organize data
pf$Population <- as.factor(pf$Nutrient)

P<-pf %>% 
  mutate(
    Food = case_when(
      grepl("MD", Population) ~ "FT1")) %>% ## FT1 is food fed in fritz lab during larval development 
  mutate( 
    Region_of_Origin = case_when(
      grepl("FL", Population) ~ "FL",
      grepl("ME", Population) ~ "ME",
      grepl("AG", Population) ~ "AG",
      grepl("BG", Population) ~ "BG",
      grepl("TX", Population) ~ "TX"))

P$Food<-as.factor(P$Food)  
P$Region_of_Origin<-as.factor(P$Region_of_Origin) 
P$Concentration<-as.numeric(P$Concentration) 

P %>% filter(Food == 'FT1') -> P_MD  ###switch Food == '' between FT1 and FT2 to get the one you want ###

##Create violin plot w/bargraph
protein <- ggplot(P_MD, aes(x = Region_of_Origin, y = Concentration, fill = Nutrient)) +
  geom_violin(trim = TRUE) +
  labs(x = "Population", y = "Amount Protein per Weight per Individual Larvae (μg/mg)", fill = "Nutrient") +
  scale_fill_manual(values= color_order) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  geom_boxplot(width=0.1) + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
protein


```



# Glycogen


```{r, echo = FALSE}
### Glycogen ##########################################################
######################################################################

##Load data
gf <- readxl::read_xlsx("Culex_Glycogen_July2025.xlsx")

##Organize data
gf$Population <- as.factor(gf$Nutrient)

G<-gf %>% 
  mutate(
    Food = case_when(
      grepl("MD", Population) ~ "FT1")) %>%
  mutate( 
    Region_of_Origin = case_when(
      grepl("FL", Population) ~ "FL",
      grepl("ME", Population) ~ "ME",
      grepl("AG", Population) ~ "AG",
      grepl("BG", Population) ~ "BG",
      grepl("TX", Population) ~ "TX"))

G$Food<-as.factor(G$Food)  
G$Region_of_Origin<-as.factor(G$Region_of_Origin) 
G$Concentration<-as.numeric(G$Concentration) 

G %>% filter(Food == 'FT1') -> G_MD


##Create violin plot w/bargraph
g<-ggplot(G_MD, aes(x = Region_of_Origin, y = Concentration, fill = Nutrient)) +
  geom_violin(trim = TRUE) +
  labs(x = "Population", y = "Amount Glycogen per Weight per Individual Larve (μg/mg)", fill = "Nutrient") +
  scale_fill_manual(values= color_order) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  geom_boxplot(width=0.1) + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
g

```



# Total Lipids



**FT1**

```{r, echo = FALSE}
### Total Lipids #################################################################
##################################################################################

##Load data
lf <- readxl::read_xlsx("Culex_Total Lipids_July2025.xlsx")

##Organize data
lf$Population <- as.factor(lf$Nutrient)

L<-lf %>% 
  mutate(
    Food = case_when(
      grepl("MD", Population) ~ "FT1")) %>%
  mutate( 
    Region_of_Origin = case_when(
      grepl("FL", Population) ~ "FL",
      grepl("ME", Population) ~ "ME",
      grepl("AG", Population) ~ "AG",
      grepl("BG", Population) ~ "BG",
      grepl("TX", Population) ~ "TX"))

L$Food<-as.factor(L$Food)  
L$Region_of_Origin<-as.factor(L$Region_of_Origin) 
L$Concentration<-as.numeric(L$Concentration) 

L %>% filter(Food == 'FT1') -> L_MD

##Create violin plot w/bargraph
l<-ggplot(L_MD, aes(x = Region_of_Origin, y = Concentration, fill = Nutrient)) +
  geom_violin(trim = TRUE) +
  labs(x = "Population", y = "Amount Total Lipids per Weight per Individual Larvae (μg/mg)", fill = "Nutrient") +
  scale_fill_manual(values= color_order) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  geom_boxplot(width=0.1) + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
l

```


# Triglycerides


```{r, echo = FALSE}
### Triglycerides #################################################################
###################################################################################

##Load data
tf <- readxl::read_xlsx("Culex_Triglycerides_July2025.xlsx")

##Organize data
tf$Population <- as.factor(tf$Nutrient)

Tf<-tf %>% 
  mutate(
    Food = case_when(
      grepl("MD", Population) ~ "FT1")) %>%
  mutate( 
    Region_of_Origin = case_when(
      grepl("FL", Population) ~ "FL",
      grepl("ME", Population) ~ "ME",
      grepl("AG", Population) ~ "AG",
      grepl("BG", Population) ~ "BG",
      grepl("TX", Population) ~ "TX"))

Tf$Food<-as.factor(Tf$Food)  
Tf$Region_of_Origin<-as.factor(Tf$Region_of_Origin) 
Tf$Concentration<-as.numeric(Tf$Concentration)

Tf %>% filter(Food == 'FT1') -> Tf_MD

t<-ggplot(Tf_MD, aes(x = Region_of_Origin, y = Concentration, fill = Nutrient)) +
  geom_violin(trim = TRUE) +
  labs(x = "Population", y = "Amount Triglycerides per Weight per Individual Larvae (μg/mg)", fill = "Nutrient") +
  scale_fill_manual(values= color_order) +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  geom_boxplot(width=0.1) + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
t

```


